A common misapplication of statistical inference: nuisance control with null-hypothesis significance test

### 2016-05-20 response.md

# Response to reviewers

We thank both reviewers for the insightful discussions.
Beyond their overall encouragement, they have identified terminological and conceptual errors on our parts and given additional methodological advice for which we are very grateful.
We note the reviewers were largely in agreement with regards to their overall evaluation and particular criticisms.
We have found it straight-forward to include all of their advice, and think this has greatly improved the manuscript.

In particular, reviewer #1 has pointed out that we employed the term parameter not only in the sense of e.g. a population parameter with which inference is concerned, but also incorrectly to refer to (confounding) covariates ("nuisance parameter").
In the reworked version of the manuscript, the term parameter is only employed in the sense of a population parameter to be estimated.
We have also corrected the other confusingly written or wrong uses of statistical terminology and concepts, such as those concerning variance in parameter estimation via multiple regression (which reviewer #2 has also pointed out).

@@

Reviewer #2 (Jake Westfall) inquires about quantitative estimates of the failure rates of this approach.
We have amended the methods section with a short simulation, where we show that in the described procedure, in realistic contexts, failures (e.g. failures to reject confounded samples when a significant effect is only observed due to confounding) occur more often than correct rejections (e.g. correctly rejecting a sample for confound imbalance).

We have also rephrased multiple sections, shortened the discussion of collinearity, extended the discussions of the error of "accepting" the null, further emphasized the importance of providing descriptive statistics (which both reviewers have repeatedly asked for), reworked and shortened the confusingly written section on employing multiple regression to control for confounds (including a reference to Westfall & Yarkoni, 2016, which appeared just days after our initial submission).
Finally, we have included a simulation that demonstrates the impact of the problem.

### 2016-05-20 response.md

# Response to reviewers

We thank both reviewers for the insightful discussions.
Beyond their overall encouragement, they have identified terminological and conceptual errors on our parts and given additional methodological advice for which we are very grateful.
We note the reviewers were largely in agreement with regards to their overall evaluation and particular criticisms.
We have found it straight-forward to include all of their advice, and think this has greatly improved the manuscript.

In particular, reviewer #1 has pointed out that we employed the term parameter not only in the sense of e.g. a population parameter with which inference is concerned, but also incorrectly to refer to (confounding) covariates ("nuisance parameter").
In the reworked version of the manuscript, the term parameter is only employed in the sense of a population parameter to be estimated. We instead employ the term "nuisance variable" to refer to that which we had previously incorrectly called "nuisance parameter".
We have also corrected the other confusingly written or wrong uses of statistical terminology and concepts, such as those concerning variance in parameter estimation via multiple regression (which reviewer #2 has also pointed out).

@@

Reviewer #2 (Jake Westfall) inquires about quantitative estimates of the failure rates of this approach.
We have amended the methods section with a short simulation, where we show that in the described procedure, in realistic contexts, failures (e.g. failures to reject confounded samples when a significant effect is only observed due to confounding) occur more often than correct rejections (e.g. correctly rejecting a sample for confound imbalance).

We have also rephrased multiple sections for style, shortened the discussion of collinearity, extended the discussions of the error of "accepting" the null, further emphasized the importance of providing descriptive statistics (which both reviewers have repeatedly asked for), reworked and shortened the confusingly written section on employing multiple regression to control for confounds (including a reference to Westfall & Yarkoni, 2016, which appeared just days after our initial submission).
Finally, we have included a simulation that demonstrates the impact of the problem.

We again wish to thank the reviewers for their fair and constructive criticism.

### 2016-07-06 simulation/statfail.Rmd
missing space
slight improvements in graphics, hopefully clearer phrasing (that also matches new phrasing in paper)

---
title: 'Inference failure: Confound and control in BBS experiments'
author: "Phillip M. Alday"
date: "May  2016"
output: 
  html_document:
    toc: TRUE
    toc_float: TRUE
    code_folding: hide
    self_contained: TRUE
    highlight: textmate
params:
  n: 500
  use_plotly: FALSE
  cores: 2
---

@@

Another Python-based simulation will soon be made available on Binder, but the preliminary code is already in the GitHub repository.

@@

1. **manipulation effect size:** the effect size from the manipulation of interest
2. **confound feature size:** the measured size of the confounding feature
3. **confound feature-effect correlation:**the correlation between that measured size the and the effect size of the feature

@@

@@

The difficult part of the simulation is performed by `statfail.R`, which is shared with the Shiny app.
Both the source code and an explanation of the general principles (`simulation_notes.md`) are available in the GitHub repository.
Here, we only need to iterate over a few possibilities for the manipulation effect size, confound feature size and confound feature-effect correlation.
The number of simulated experiments is held constant for all combinations of these (set via the document parameter `n`, which had the value `r params$n` when this document was generated).
Increasing this value will of course help detect rare phenomena, but will drastically increase time and memory needed.

@@

## Generate graphics
```{r generate.graphics, dependson='simulate',cache=TRUE,message=FALSE}
# prettier names for the legend
results.long$type <-factor(results.long$type
                           ,levels=c("rejections"
                                     ,"manipulation_still_significant"
                                     ,"feature_has_no_effect"
                                     ,"feature_irrelevant_in_multiple"
                                     ,"not_rejected"
                                     ,"feature_relevant_in_multiple"
                                     ,"manipulation_only_in_simple")
                           ,labels = c("rejected"
                                       ,"but significant manipulation missed"
                                       ,"when feature was not significant in simple regression"
                                       ,"when feature was not significant in multiple regression"
                                       ,"accepted"
                                       ,"but feature still had an effect in multiple regression",
                                       "and the manipulation was only significant in simple regression"))

# This uses the deprecated labeller API for ggplot.
facet_labeller <- function(variable,value){
  lbl <- NULL
  if (variable=='cfs') {
    lbl <- paste("Feature:",value)
  }else if (variable == 'mes') {
    lbl <- paste("Manipulation:",value)
  }else if (variable == 'cfec') {
    lbl <- paste("Feature-size/effect correlation:",value)
  } else {
    lbl <- as.character(value)
  }
}

# generate the graphic for rejections
gg.reject <- ggplot(subset(results.long,result=="reject"),aes(color=type,x=cfec,y=percent.studies)) +
  geom_line(stat="identity") + 
  facet_grid(cfs ~ mes,labeller = facet_labeller) + 
  xlab("Correlation of feature as measured and effect of feature on outcome") + 
  ylab("Percent of simulated studies") + 
  theme(legend.position="top") + #coord_fixed(ratio=1/100) +
  guides(color=guide_legend(title="Result based on inferential pretesting of measured feature",
                            title.position="top",
                            nrow=2))

@@

## Acceptances
* Acceptances where the feature still had an effect in multiple rejection are failed detections of a real confounding effect.
* Acceptances where the manipulation was only significant in simple regression are failed detections of a real confounding effect *that completes subsumes the manipulation effect*.

>>>>

---
title: 'A common misapplication of statistical inference: nuisance control with null-hypothesis significance test'
author: "Phillip M. Alday"
date: "May  2016"
output: 
  html_document:
    toc: TRUE
    toc_float: TRUE
    code_folding: hide
    self_contained: TRUE
    highlight: textmate
params:
  n: 500
  use_plotly: FALSE
  cores: 2
---

@@

<!--Another Python-based simulation will soon be made available on Binder, but the preliminary code is already in the GitHub repository.-->

@@

1. **manipulation effect size:** the effect size from the manipulation of interest (given as Cohen's $d$ between groups)
2. **confound size:** the measured size of the confounding feature (given as Cohen's $d$ between groups)
3. **confound-outcome correlation:** the simple correlation between that the confound size and the outcome (given as Pearson's $r$)

@@

In other words, the confound-outcome correlation is the actual effect size of the confounding variable, taken together with the confound size, you can compute the actual impact of confounding on the manipulation.

@@

The difficult part of the simulation is performed by `statfail.R`, which is shared with the Shiny app.
Both the source code and an explanation of the general principles (`simulation_notes.md`) are available in the GitHub repository.
Here, we only need to iterate over a few possibilities for the manipulation effect size, confound size, and confound-outcome  correlation.
The number of simulated experiments is held constant for all combinations of these (set via the document parameter `n`, which had the value `r params$n` when this document was generated).
Increasing this value will of course help detect rare phenomena, but will drastically increase time and memory needed.

@@

## Generate graphics
```{r generate.graphics, dependson='simulate',cache=TRUE,message=FALSE,warning=FALSE}
# prettier names for the legend
results.long$type <-factor(results.long$type
                           ,levels=c("rejections"
                                     ,"manipulation_still_significant"
                                     ,"feature_has_no_effect"
                                     ,"feature_irrelevant_in_multiple"
                                     ,"not_rejected"
                                     ,"feature_relevant_in_multiple"
                                     ,"manipulation_only_in_simple")
                           ,labels = c("rejected"
                                       ,"but significant manipulation missed"
                                       ,"when feature was not significant in simple regression"
                                       ,"when feature was not significant in multiple regression"
                                       ,"accepted"
                                       ,"but feature still had an effect in multiple regression",
                                       "and the manipulation was only significant in simple regression"))

# This uses the deprecated labeller API for ggplot.
facet_labeller <- function(variable,value){
  lbl <- NULL
  if (variable=='cfs') {
    lbl <- paste("Feature:",value)
  }else if (variable == 'mes') {
    lbl <- paste("Manipulation:",value)
  }else if (variable == 'cfec') {
    lbl <- paste("Feature-outcome correlation",value)
  } else {
    lbl <- as.character(value)
  }
}

# generate the graphic for rejections
gg.reject <- ggplot(subset(results.long,result=="reject"),aes(color=type,x=cfec,y=percent.studies)) +
  geom_line(stat="identity") + 
  facet_grid(cfs ~ mes,labeller = facet_labeller) + 
  xlab("Correlation of confounding feature and dependent variable") + 
  ylab("Percent of simulated studies") + 
  theme(legend.position="top", axis.text.x = element_text(angle = 45, hjust = 1), strip.text.y = element_text(angle=0)) + #coord_fixed(ratio=1/100) +
  guides(color=guide_legend(title="Result based on inferential pretesting of confounding feature",
                            title.position="top",
                            nrow=2))

@@

## Acceptances
* Acceptances where the feature still had an effect in multiple rejection are failed detections of a real confound.
* Acceptances where the manipulation was only significant in simple regression are failed detections of a real confound *that completely subsumes the manipulation effect*.
